---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
opioid = read_csv("Medicaid_Opioid_Prescribing_Rates_by_Geography_2020")
```

# Data 1

## Introduction and data

-   The data is exported from the Centers for Medicare & Medicaid Services.

-   The data was last collected in 2020 from CMS administrative data for beneficiaries enrolled in Medicaid available from the Medicaid State Drug Utilization program.

-   The Medicaid Prescribing Opioid Rates data set provides an overview of the geographic comparisons of opioid prescribing rates by state, county, and zip code. Each observation provides information on geographical areas(country, state, county, and zip code) in the United States and their correlation with current and long-acting opioid rates and medicaid claims.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    How do opioid prescription rates differ by geography and time?

    How are opioid prescription rates expected to change in the future based on geography?

    Is an individual's medicaid plan predictive of their geography and opioid insurance claim?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    This research topic will explore the correlation between geography, insurance plans, and the opioid epidemic. I would expect low-socioeconomic regions to be prone to high opioid prescriptions as the majority of opioid users are low income individuals. Also, I would expect the more generous insurance plans to have high variability with opioid prescribing rates as moral hazard contributes to accepting prescriptions.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    Data Year- Character

    geographical level- categorical

    geographical code- categorical

    geographical description- categorical

    plan type- categorical

    opioid claims- quantitative

    overall claims- quantitative

    opioid prescribing rate- quantitative

    five year change in opioid prescribing rate- quantitative

    one year change in opioid prescribing rate- quantitative

    long-acting opioid claims- quantitative

    long-acting prescribing rate- quantitative

    five year change in long acting opioid prescribing rate- quantitative

    one year change in long-acting opioid prescribing rate- quantitative

-   

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
```

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

# add code here
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
```
