---
title: "The Wranglers Project"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
opioid <- read_csv("data/Medicaid_Opioid_2020.csv")
```

# Data 1 (Opioids)

## Introduction and data

-   Identify the source of the data.

    The data is exported from the Centers for Medicare & Medicaid Services.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    The data was last collected in 2020 from CMS administrative data for Medicaid beneficiaries available from the Medicaid State Drug Utilization program.

-   Write a brief description of the observations.

    The Medicaid Prescribing Opioid Rates data set provides an overview of the geographic comparisons of opioid prescribing rates by state, county, and zip code. Each observation provides information on geographical areas(country, state, county, and zip code) in the United States and their correlation with current and long-acting opioid rates and medicaid claims from 2016 to 2020.

## Research question

-   A well formulated research question.

    How do opioid prescription rates differ by geography and time? How are opioid prescription rates expected to change in the future based on geography? Is an individual's medicaid plan predictive of their geography and opioid insurance claim?

    These questions are important because their answers provide in depth information into the roots of the opioid epidemic and possible solutions.

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    This research topic will explore the correlation between geography, insurance plans, and the opioid epidemic. I would expect low-socioeconomic regions to be prone to high opioid prescriptions as the majority of opioid users are low income individuals. Also, I would expect the more generous insurance plans to have high variability with opioid prescribing rates as moral hazard contributes to accepting prescriptions.

    This date set raises ethical concerns relating to the accuracy of the prescription rate reportings. Because of the controversial topic of drugs, many hospitals and physicians may misreport opioid prescription rates. Also, this data set does not consider opioids that were illegally prescribed and bought. Therefore, this data set may not provide our team with all of the information regarding opioid use in the US.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    Data Year - character

    geographical level - categorical

    geographical code - categorical

    geographical description - categorical

    plan type - categorical

    opioid claims - quantitative

    overall claims - quantitative

    opioid prescribing rate - quantitative

    five year change in opioid prescribing rate - quantitative

    one year change in opioid prescribing rate - quantitative

    long-acting opioid claims - quantitative

    long-acting prescribing rate - quantitative

    five year change in long acting opioid prescribing rate - quantitative

    one year change in long-acting opioid prescribing rate - quantitative

## Literature

-   Find one published credible article on the topic you are interested in researching.

    "Geographic Variation in Opioid Prescribing in the US" by Douglas C. McDonald

    https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3509148/

-   Provide a one paragraph summary about the article.

    This article examines the relationship between geographic locations within the United States and rates of opioid prescribing aiming to discover connections between the opioid epidemic and geographic location. The study found that poorer areas, especially Western Appalachia, have significantly higher rates of opioid prescribing. Other factors that the study considered include the availability of physicians, the general strength of healthcare resources in the area, prescription monitoring laws, characteristics of the area's populations, and other factors. The study found that the emergence of higher demand for treatment with opioids coupled with the exhaustion of medical resources along with the weak medical consensus on the appropriate use of opioids factor into the over-prescription crisis fueling the opioid epidemic.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    Our research question builds on the article cited by limiting the scope of patients to only Medicare beneficiaries. Because our data set only includes Medicare patients, our team can analyze how opioids are being prescribed to low-income and/or disabled individuals in the United States.

## Glimpse of data

```{r}
#| label: load-data-1

glimpse(opioid)
```

# Data 2 (Police Shootings)

## Introduction and data

-   Identify the source of the data.

    This data is from the CORGIS Dataset Project and was created in 9/15/2021. The data was compiled from the Washington Post.

<!-- -->

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    The Washington Post regularly collects this data as police shootings happen. The Post collected the data by looking at local news reports, law enforcement websites and social media, and by monitoring independent databases such as Killed by Police and Fatal Encounters. In necessary cases, the Post did additional research.

<!-- -->

-   Write a brief description of the observations.

    This dataset collects data on fatal shootings in the United States by a police officer in the line of duty from Jan. 1, 2015 to Sept. 11, 2021. This data includes gender, race, mental illness, location, and date. Each observation represents a fatal shooting event by a police officer.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    How do race, gender, and mental illness affect rates of police shootings over time across the United States?

<!-- -->

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    Our dataset includes 6570 rows meaning that at least 6570 people were killed by police from 2015 to 2021. By discovering how different factors affect the number of shootings, we can begin to understand why these tragedies happen, and what we can do to stop them from happening.

    The research topic looks at the correlation between race, gender, and mental illness and police shooting rates. We would expect that male individuals of minority communities with previous mental illnesses would be more likely to be affected.

    One ethical concern is the possibility of misinformation in the data and a lack of accurate reporting. Another concern is consent and whether the victims families want their loved ones to have identifying information (like their name and information about their death) in a public database.

<!-- -->

-   Identify the types of variables in your research question. Categorical? Quantitative?

    Person.Name - String\
    Person.Age - Integer Person.Gender - String\
    Person.Race - String\
    Incident.Date.Month - Integer Incident.Date.Day - Integer\
    Incident.Date.Year - Integer\
    Incident.Date.Full - String\
    Incident.Location.City - String Incident.Location.State - String\
    Factors.Armed - String\
    Factors.Mental-Illness - Boolean\
    Factors.Threat-Level - String\
    Factors.Fleeing - String\
    Shooting.Manner - String\
    Shooting.Body-Camera - Boolean

## Literature

-   Find one published credible article on the topic you are interested in researching.

    "What the data say about police shootings" by Lynne Peeples

    https://www.nature.com/articles/d41586-019-02601-9

-   Provide a one paragraph summary about the article.

    Police officers in the United States fatally shoot about three people per day on average, a number that is close to the yearly totals for other wealthy nations. Data on these deadly encounters has been hard to come by, but five years later, newspapers, enterprising individuals and the federal government have launched ambitious data-collection projects to fill the gaps and improve transparency and accountability over how police officers exercise their right to use deadly force. Social scientists and public-health researchers have produced more than 50 publications to address pressing questions, such as whether the police are disproportionately quick to shoot black civilians and those from other minority groups.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    New databases show that police officers' use of lethal force is much more common than previously thought, and varies significantly across the country. We plan to track these differences which could help identify opportunities to reduce the number of shootings and deaths for both civilians and police officers.

## Glimpse of data

```{r}
#| label: load-data-2

library(readr)
police_shootings <- read_csv("data/police_shootings.csv")
glimpse(police_shootings)
```

# Data 3 (SAT Scores)

## Introduction and data

-   Identify the source of the data.

    We obtained this dataset from the CORGIS Dataset Project.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    It was collected by Dr. Austin Cory Bart (currently an Assistant Professor at the University of Delaware) in October, 2016 from the Common Core of Data (CCD), an organization that aims to "provide basic information on public elementary and secondary schools" within the United States.

-   Write a brief description of the observations.

    Each observation represents a state's annual SAT score report and demographic information.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    How do academic history and demographic factors (like family income and gender) affect standardized test scores of high school students in the United States? How should this inform our interpretation of SAT results?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    For this research project, we hope to investigate education inequities in the United States by examining the academic and social factors that influence student performance on the SAT. We hypothesize that social and demographic factors will be a better predictor of SAT score results than academic history.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    Year - Integer

    State.Code - String

    Total.Math - Integer

    Total.Verbal - Integer

    Total.Test-takers - Integer

    Academic Subjects.English.Average GPA - Float (available for several subject areas)

    Academic Subjects.English. Average Years - Float (available for several subject areas)

    Family Income.Between 20-40k.Test-takers - Integer (available for several income brackets and test types)

    GPA.A minus.Test-takers - Integer (available for several GPA levels and test types)

    Gender.Female.Test-takers - Integer (available for male and female students and several test types)

    Score Ranges - Integer (available for several test sections)

## Literature

-   Find one published credible article on the topic you are interested in researching.

    "Is Income Implicit in Measures of Student Ability?" by Jason Sockin

    https://budgetmodel.wharton.upenn.edu/issues/2021/9/28/is-income-implicit-in-measures-of-student-ability

-   Provide a one paragraph summary about the article.

    SAT and other standardized tests have become a heavily debated aspect of student education and college admissions, as many US universities have decided to opt for a test-optional admissions process. This brings attention to how possible socioeconomic factors affect student's academic record, including GPA, high school rank, and standardized test scores. According to data from the National Longitudinal Survey of Youth 1997 analyzed in this article, it was found that income had a weaker correlation with GPA and class rank than the correlation between income and SAT score. Similar correlation strengths were observed for ACT data as well. This article concludes that standardized testing reveals some abilities that are not tested in school and are impacted by environmental factors.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    Building upon this article, our research question will utilize a more current dataset that was compiled in 2016. Additionally, our research question will explore more environmental factors and context, like location and gender.

## Glimpse of data

```{r}
#| label: load-data-3

library(readr)
school_SAT <- read_csv("data/school_scores.csv")
glimpse(school_SAT)
```
