---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   The data is exported from the Centers for Medicare & Medicaid Services.

-   The data was last collected in 2020 from CMS administrative data for beneficiaries enrolled in Medicaid available from the Medicaid State Drug Utilization program.

-   The Medicaid Prescribing Opioid Rates data set provides an overview of the geographic comparisons of opioid prescribing rates by state, county, and zip code. Each observation provides information on geographical areas(country, state, county, and zip code) in the United States and their correlation with current and long-acting opioid rates and medicaid claims.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    How do opioid prescription rates differ by geography and time?

    How are opioid prescription rates expected to change in the future based on geography?

    Is an individual's medicaid plan predictive of their geography and opioid insurance claim?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    This research topic will explore the correlation between geography, insurance plans, and the opioid epidemic. I would expect low-socioeconomic regions to be prone to high opioid prescriptions as the majority of opioid users are low income individuals. Also, I would expect the more generous insurance plans to have high variability with opioid prescribing rates as moral hazard contributes to accepting prescriptions.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    Data Year- Character

    geographical level- categorical

    geographical code- categorical

    geographical description- categorical

    plan type- categorical

    opioid claims- quantitative

    overall claims- quantitative

    opioid prescribing rate- quantitative

    five year change in opioid prescribing rate- quantitative

    one year change in opioid prescribing rate- quantitative

    long-acting opioid claims- quantitative

    long-acting prescribing rate- quantitative

    five year change in long acting opioid prescribing rate- quantitative

    one year change in long-acting opioid prescribing rate- quantitative

-   

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
```

# Data 2

## Introduction and data

This data is from the CORGIS Dataset Project. The data was taken from the Washington Post. The Post collected the data by looking at local news reports, law enforcement websites and social media, and by monitoring independent databases such as Killed by Police and Fatal Encounters. In necessary cases, the Post did additional research. 

The data was put together on 9/15/2021. The Post regularly collects this data as the shootings happen. 


This dataset collects data on fatal shootings in the United States by a police officer in the line of duty from Jan. 1, 2015 to Sept. 11, 2021. This data includes gender, race, mental illness, location, and date.


## Research question

How do race, gender, and mental illness affect rates of police shootings over time across the United States?

Our dataset includes 6570 rows meaning that at least 6570 people were killed by police from 2015 to 2021. By discovering how different factors affect the number of shootings, we can begin to understand why these tragedies happen, and what we can do to stop them from happening.

The research topic looks at the correlation between race, gender, and mental illness and police shooting rates. We would expect that male individuals of minority communities with previous mental illnesses would be more likely to be affected. 

Person.Name	- String	
Person.Age -	Integer
Person.Gender -	String	
Person.Race	- String	
Incident.Date.Month	- Integer
Incident.Date.Day	- Integer	
Incident.Date.Year	- Integer	
Incident.Date.Full	- String	
Incident.Location.City	- String
Incident.Location.State	- String	
Factors.Armed	- String	
Factors.Mental-Illness	- Boolean	
Factors.Threat-Level -	String	
Factors.Fleeing	- String	
Shooting.Manner	- String	
Shooting.Body-Camera	- Boolean	

## Literature

Link to article.
https://www.nature.com/articles/d41586-019-02601-9 

Police officers in the United States fatally shoot about three people per day on average, a number that is close to the yearly totals for other wealthy nations. Data on these deadly encounters has been hard to come by, but five years later, newspapers, enterprising individuals and the federal government have launched ambitious data-collection projects to fill the gaps and improve transparency and accountability over how police officers exercise their right to use deadly force. Social scientists and public-health researchers have produced more than 50 publications to address pressing questions, such as whether the police are disproportionately quick to shoot black civilians and those from other minority groups. New databases show that police officers' use of lethal force is much more common than previously thought, and varies significantly across the country. 

We plan to track these differences which could help identify opportunities to reduce the number of shootings and deaths for both civilians and police officers.


## Glimpse of data

```{r}
#| label: load-data-2
library(readr)
police_shootings <- read_csv("data/police_shootings.csv")
glimpse(police_shootings)
# add code here
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
```
