---
title: "An Analysis of Police Shootings in the United States from 2015-2021"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: false
---

```{r}
#| label: load-data-2
library(tidyverse)
library(tidymodels)
library(readr)
library(infer)
police_shootings <- read_csv("data/police_shootings.csv")
glimpse(police_shootings)
```

**Introduction**

Created on September 15, 2021, the `police_shootings` data set is sourced from the CORGIS Dataset Project and was compiled from findings by *The Washington Post*. *The Washington Post* regularly collects this data as police shootings happen by monitoring local news reports, law enforcement websites, social media, and independent databases such as Killed by Police and Fatal Encounters. When necessary, *The Post* did additional research to investigate each individual case.

This data set describes fatal shootings in the United States by a police officer in the line of duty from January 1, 2015, to September 11, 2021. Each observation represents a fatal shooting event by a police officer. This data includes information on the victim's gender, race, mental illness, location, and date of the event.

In a [*Nature* article](https://www.nature.com/articles/d41586-019-02601-9) titled "What the data say about police shootings," Lynne Peeples explains that police officers in the United States fatally shoot about three people per day on average, a number that rivals the yearly totals of other wealthy nations. Data on these deadly encounters had been hard to come by, but five years later, newspapers, enterprising individuals, and the federal government have launched ambitious data-collection projects to fill the gaps and improve transparency and accountability over how police officers exercise their right to use deadly force. Social scientists and public health researchers have produced more than 50 publications to address pressing questions, such as whether the police are disproportionately quick to shoot black civilians and those from other minority groups. This topic is extremely important to investigate because it brings awareness to the implicit biases that the US police force may hold. 

**Research question**

How do race, gender, age, and whether a civilian was armed or not affect rates of police shootings across the United States from 2015 to 2021?

**Data**

Our data set includes 6569 rows, indicating that at least 6569 people were killed by police from 2015 to 2021. By discovering how different factors affect the number of shootings, we can begin to understand why these tragedies happen and what we can do to stop them from happening.

Our research topic examines police shooting history on the basis of race, gender, and age. We hypothesize that male individuals who identify as a POC are perceived as an increase threat during fatal police shootings.

**Ethical Concerns**

One ethical concern is the possibility of misinformation in the data and a lack of accurate reporting. Not every police shooting is reported or handled correctly. Although *The Washington Post* collected the data from many different sources, there is no way of guaranteeing that the data set is fully comprehensive.

Second, as demonstrated below, all 6,569 victims identified in this data set have been characterized as having a mental illness. As a result, we decide not to make a visualization based on the mental illness variable. We believe this would pose an ethical issue because mental health is a spectrum with varying characteristics and severity of impact. Having a mental illness cannot be reduced to a simple "TRUE" or "FALSE." Instead, we decided that other variables would be a better focus.

```{r}
#| label: mental-illness

sum(police_shootings$`Factors.Mental-Illness`)
```

Finally, a third concern is consent and whether the victims families want their loved ones to have identifying information (like name and information about their death) in a public database. This is a highly sensitive topic and should be treated as such.

For the purposes of this project, we will assume that the following data was collected randomly and that observations are independent.

We have identified several key variables that will advance our research question: 

`Person.Age` - An integer value of the victim's age

`Person.Gender` - The victim's identified gender

`Person.Race` - The victim's identified race

`Incident.Location.State` - The state in which the shooting incident happened 

`Factors.Armed` - Indicates whether the victim was armed

`Factors.Threat-Level` - The level of threat that the victim was perceived by the police (laveled as "attack", "undetermined", "other", or "unknown")


**Methodology**

The average age for a police shooting victim is 35; whereas, the median age is 34. We can conclude that the data is not central to 34-35 years of age as the standard deviation of the the victim's age is \~15 years.

```{r}
#| label: summary-stats

police_shootings |>
  summarize(mean_age = mean(Person.Age),
            median_age = median(Person.Age),
            stdv_age = sd(Person.Age))
```

```{r}
police_shootings |>
  mutate(state = Incident.Location.State)
```

```{r}
#| label: usmap-shootings
library(usmap) 
library(ggplot2) 

data2 <- police_shootings |> 
  mutate(state = Incident.Location.State) |>
  group_by(state) |> 
  count() 

plot_usmap(data = data2, values = "n", regions = "states", color = "red") + 
  scale_fill_continuous(low = "white", high = "red", 
                        name = "Frequency of 
                                \nPolice Shootings") +
  theme(legend.position = c(0.9, 0.3),
        legend.key.size = unit(.25, "cm")) +
  labs(title = "Location of Fatal Police Shootings by State",
      subtitle = "2015-2021")
```

As seen by the map, the majority of our data was collected in California. This could be a limitation to our data as it demonstartes biases.

In yearly-analysis, we chose to do a line graph to represent the progression of shooting per day over time as it effectively reveals trends and changes in shooting per day across the years.

```{r}
#| label: yearly-analysis

yearly_shootings <- police_shootings |>
  group_by(Incident.Date.Year) |>
  count(Incident.Date.Year) 

days_in_year <- c(365, 366, 365, 365, 365, 366, 253)

yearly_shootings |>
  ggplot(aes(x = Incident.Date.Year, y = n/days_in_year)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) + 
  geom_line() +
  labs(
    title = "Fig. 1: Shootings per Day from 2015 to 2021",
    x = "Year", 
    y = "Number of Shootings per Day") +
  theme_minimal()
```

As seen by the graph below, \~10% of male victims and 13% of female victims were unarmed.

```{r}
#| label: arms-status

police_shootings |>
  filter(Factors.Armed != "unknown") |>
  mutate(Factors.Armed.Test = if_else(
    Factors.Armed == "unarmed" | Factors.Armed == "toy weapon", "Unarmed", 
    "Armed with weapon")) |>
  ggplot(aes(x = Person.Gender, fill = Factors.Armed.Test)) +
  geom_bar(position = "fill") +
  labs(
    title = "Fig. 2: Arms Status of Victims by Gender",
    y = "Percentage of Victims",
    x = "Sex",
    fill = "Arms Status") +
  theme_minimal()
```

We used box plots to visualize male versus female and race as it is the best way to show distributions of numeric data values between multiple groups. It made the differences in age between male and female victims and victims of different races very clear. We could compare the median, range, and outliers.

```{r}
#| label: box-plots-race

police_shootings |> 
  filter(Person.Race != "unknown") |>
  ggplot(
    aes(x = Person.Age, y = Person.Race)) + 
    geom_boxplot() + 
    labs(title = "Fig. 3: Race and Age of Victims",
         x = "Age (years)",
         y = "Race",
         color = "Race") +
  theme_minimal()
```

```{r}
#| label: box-plot-gender

police_shootings |> 
  filter(Person.Gender != "Unknown") |>
  ggplot(
    aes(x = Person.Age, y = Person.Gender)) + 
    geom_boxplot() + 
    labs(title = "Fig. 4: Sex and Age of Victims",
         x = "Age (years)",
         y = "Sex",
         color = "Sex") +
  theme_minimal()
```

```{r}
#| label: bar-graph-race

total <- police_shootings |> 
  summarize(total = n()) |>
  pull(total)

race_prop <- police_shootings |> 
  group_by(Person.Race) |>
  summarize(prop = n()/total)
  
#race_prop |>
  #ggplot(aes(y = prop, x = Person.Race, stat = "count")) + 
  #  geom_bar() + 
  #  labs(title = "Fig. 5: Number of Victims by Race",
   #      x = "Race",
    #     y = "Count",
    #     fill = "Race") +
 # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        #legend.position = "none")
```

```{r}
#| label: bar-graph-arms-status-race

police_shootings |>
  filter(Factors.Armed != "unknown") |>
  mutate(Factors.Armed.Test = if_else(
    Factors.Armed == "unarmed" | Factors.Armed == "toy weapon", "Unarmed", 
    "Armed with weapon")) |>
  ggplot(aes(x = Person.Race, fill = Factors.Armed.Test)) +
  geom_bar(position = "fill") +
  labs(
    title = "Fig. 6: Arms Status of Victims by Race",
    y = "Proportion of Victims",
    x = "Race",
    fill = "Arms Status") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r}
#| label: fit-model 

police_shootings_fit <- police_shootings |>
  filter(`Factors.Threat-Level` != "unknown") |>
  mutate(Factors.Armed.Test = if_else(Factors.Armed == "unarmed" | 
                                      Factors.Armed == "toy weapon", "unarmed",
                                      "armed with weapon"),
         race = if_else(Person.Race != "White", "POC", "White"), 
        `Factors.Threat-Level`= if_else(`Factors.Threat-Level` == "attack", 
                                      "attack", "not specified as attack"),
        `Factors.Threat-Level`= as.factor(`Factors.Threat-Level`)) 

best_logreg_model <- logistic_reg() |> 
  set_engine("glm") |>
  fit(`Factors.Threat-Level` ~ race + Person.Gender + Factors.Armed.Test + Person.Age, 
      data = police_shootings_fit, family = "binomial") 

best_logreg_model |> 
  glance() |>
  pull(AIC)

best_logreg_model |>
  tidy()
```

$\log{\frac{p}{1-p}}$ = -0.229153274 -0.114483595 \* ($race\_white$) - 0.244889178 \* ($Person.Gender\_Male$) + 1.038480834 \* ($Person.Gender\_Unknown$) + 0.652516293 \* ($Factors.Armed.Test\_unarmed$) - 0.004189227 \* ($Person.Age$)

Where $p$ is the probability of a person being a perceived threat by the police, stating the incident as an "attack."

After fitting the logistic model and computing backwards elimination, we found the best model for our response variable (Factors.Threat.Level) to be an additive model between our four explanatory variables: gender, race, armed status, and age. During backwards elimination, we fit several additive and interactive models between the four variables and determined the lowest AIC value. Therefore, whether or not a victim was seen as a threat was explained with an additive interaction between gender, race, armed status, and age.

We also investigated whether the proportion of people perceived as an “attack” threat level was different for victims of color versus white victims. We hypothesized that the proportion of POC who were perceived as an "attack" by police is significantly greater than the proportion of white victims perceived in the same threat level category. To investigate this hypothesis, we performed a hypothesis test using a significance level of $\alpha = 0.05$. For testing purposes, we assumed independence within our data set's observations.

(Note: In the context of this question, we will consider someone as a "POC" if they do not identify as "White.")

$\pi_{POC}$ = the proportion of people who were perceived as an "attack" by the police out of those shot who identify as a POC.

$\pi_{White}$ = the proportion of people who were perceived as an "attack" by the police out of those shot who identify as White. 

$H_0: \pi_{POC} = \pi_{White}$

The proportion of people who were perceived as an "attack" by the police of those shot who identify as a POC is equal to the proportion of people who were perceived as an "attack" by the police of those shot who identify as white.

$H_A: \pi_{POC} > \pi_{White}$

The proportion of people who were perceived as an "attack" by the police of those shot who identify as a POC is greater than the proportion of people who were perceived as an "attack" by the police of those shot who identify as white.

```{r}
#| label: hypothesis-race

hypo_data <- police_shootings |>
  filter(Factors.Armed != "unknown", Person.Race != "Unknown") |>
  mutate(`Factors.Threat-Level`= if_else(`Factors.Threat-Level` == "attack", 
                                      "attack", "not specified as attack"),
        `Factors.Threat-Level`= as.factor(`Factors.Threat-Level`),
         race = if_else(Person.Race != "White", "POC", "White")) 

set.seed(69)

null_dist <- hypo_data |>
  specify(response = `Factors.Threat-Level`, 
          explanatory = race, success = "attack") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props" , order = c("POC", "White"))
```

```{r}
#| label: hypothesis-race-pvalues
hypo_data |>
  group_by(race, `Factors.Threat-Level`) |>
  summarize(props = n())

p_hat_POC <- 1764/(1764+1010)
p_hat_white <- 1912/(1912+955)
obs_diff_phat <- p_hat_POC - p_hat_white

visualize(null_dist) +
 shade_p_value(obs_stat = obs_diff_phat, direction = "greater") +
  labs(title = "Fig. 7: Simulated Null Distribution",
       x = "Difference in Proportions",
       y = "Count") +
  theme_minimal()

null_dist |>
  get_p_value(obs_stat = obs_diff_phat, direction = "greater")

```

**Results**

Based on our summary statistics for the entire data set, we observed that the mean and median age of shooting victims is 35.4 and 34.0 years, respectively. Moreover, the standard deviation of this age range is 14.9 years, demonstrating the wide spread of data points included. We noticed a slight decrease in shooting per day in 2021 though the total range of average shootings per day was small and somewhat insignificant (from roughly 2.45 shootings to 2.8 shootings). When looking at the arms status of victims by gender, female victims were more likely to be armed than male victims. One reason for this is that male victims could have been seen as more aggressive and therefore more likely to become a victim no matter if they were armed. The age distribution of victims does not vary much between races. When we look at the different races in the count of total victims, we can see white victims and African American victims were most common. Close to 2962 of the victims were white; whereas, 1552 victims were black. If race had no impact on probability of being a victim, the proportions of victims by race would be similar as the US population by race. White Americans represent 0.578 of the population. African Americans make up 0.121 of the population.

```{r}
police_shootings |>
  count(Person.Race)
white_american_prop <- 2962/6569
white_american_prop
african_american_prop <- 1552/6569  
african_american_prop
```

White Americans represent 0.578 of the US population while they only represent 0.45 of the victims. African Americans make up 0.121 of the population while they present 0.236 of the victims.Therefore we can see that African American were targeted at a higher rate than white Americans.

The race with the most unarmed victims was black and the race with the least unarmed victims was Native American. Visualization can provide ample information on implicit bias and how race impacts the police's approaches to interactions with different races. We also found the median age is roughly 34 years old for males and 35 years old for females. Male victims differ by their outliers, as many male victims are over the age of 75. The unknown victims appear to be significantly younger as seen by the median of \~21 years old. 


We performed a hypothesis test to determine whether the proportion of people perceived as an “attack” threat level was different for victims of color versus white victims. Our hypothesis test yields a p-value of 0.995. Utilizing a significance level of 0.05, we fail to reject the null hypothesis that the proportion of people who were perceived as an "attack" by the police of those shot who identify as a POC is equal to the proportion of people who were perceived as an "attack" by the police of those shot who identify as white. As a p-value of 0.995 is greater than 0.05, we have weak evidence to support the alternative hypothesis that the proportion of people who were perceived as an "attack" by the police of those shot who identify as a POC is greater than the proportion of people who were perceived as an "attack" by the police of those shot who identify as white.


**Discussion and Limitations**

  Our project allowed us to investigate how race, gender, age, and a civilian’s armed status affected perceived threat level in fatal police shootings across the US from 2015-2021. We hypothesized that male individuals who identify as a POC are perceived as an increased threat during fatal police shootings. Our statistical results do not support this hypothesis. After analyzing our data and results, we believe that our data is heavily impacted by bias and misconceptions, leading to unreliability in our results. Close to 2962 of the victims were white; whereas, 1552 victims were black. If race had no impact on probability of being a victim, the proportions of victims by race would be similar as the US population by race. White Americans represent 0.578 of the population. African Americans make up 0.121 of the population. White Americans represent 0.578 of the US population while they only represent 0.45 of the victims. African Americans make up 0.121 of the population while they present 0.236 of the victims.Therefore we can see that African American were targeted at a higher rate than white Americans. As seen by the map of fatal police shootings by state, the vast majority of shootings occurred in California. Therefore, the data is obviously includes location bias. This could affect our results as demographics change by state. Also, we believe that much of our data could be heavily impacted by bias, leading to unreliability. For example, every victim was labeled as mentally ill. This assumption created some concern for our team as we believed it was unlikely that every victim was mentally ill. Therefore, our results can support our specific data set but we do not believe that our results cannot be generalized upon a broader population beyond our sample. 
  Further exploration can allow us to analyze the relationship between each state’s victims and demographics. By computing statistical analysis for each individual state and analyzing the perceived threat level, we can further our results and conclude how each state relates to our hypothesis. 
Also, with more data, we can analyze how socioeconomic levels relate to perceived threat level. This extra demographic will allow us to further our conclusion and visualize patterns between an individual's socioeconomic level and their race, gender, age, and armed status. 
  To better our analysis, we believe that more reliable and broad data would be beneficial to our results and validity. Because our data was collected by the Corgis database through the Washington Post, our data could be very biased and invalid. By collecting data directly through the US or state governments, we believe our analysis will be more thorough and valid. 



