---
title: "An Analysis of Police Shootings in the United States from 2015-2021"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

```{r}
#| label: load-data-2
library(tidyverse)
library(readr)
library(infer)
police_shootings <- read_csv("data/police_shootings.csv")
glimpse(police_shootings)
```

**Introduction**

Created on September 15, 2021, the `police_shootings` data set is sourced from the CORGIS Dataset Project and was compiled from findings by *The Washington Post*. *The Washington Post* regularly collects this data as police shootings happen by monitoring local news reports, law enforcement websites, social media, and independent databases such as Killed by Police and Fatal Encounters. When necessary, *The Post* did additional research to investigate each individual case.

This data set describes fatal shootings in the United States by a police officer in the line of duty from January 1, 2015, to September 11, 2021. Each observation represents a fatal shooting event by a police officer. This data includes information on the victim's gender, race, mental illness, location, and date of the event.

In a [*Nature* article](https://www.nature.com/articles/d41586-019-02601-9) titled "What the data say about police shootings," Lynne Peeples explains that police officers in the United States fatally shoot about three people per day on average, a number that rivals the yearly totals of other wealthy nations. Data on these deadly encounters had been hard to come by, but five years later, newspapers, enterprising individuals, and the federal government have launched ambitious data-collection projects to fill the gaps and improve transparency and accountability over how police officers exercise their right to use deadly force. Social scientists and public health researchers have produced more than 50 publications to address pressing questions, such as whether the police are disproportionately quick to shoot black civilians and those from other minority groups.

    New databases show that police officers' use of lethal force is much more common than previously thought and varies significantly across the country. We plan to track these differences to help identify opportunities to reduce the number of police shootings and deaths for both civilians and police officers.

**Research question**

How do race, gender, and whether a civilian was armed or not affect rates of police shootings across the United States from 2015 to 2021?

**Data**

Our data set includes 6569 rows, indicating that at least 6569 people were killed by police from 2015 to 2021. By discovering how different factors affect the number of shootings, we can begin to understand why these tragedies happen and what we can do to stop them from happening.

Our research topic examines police shooting history on the basis of race, gender, and age. We hypothesize that male individuals of minority communities are most likely to consist of the largest victim population.

**Ethical Concerns**

One ethical concern is the possibility of misinformation in the data and a lack of accurate reporting. Not every police shooting is reported or handled correctly. Although *The Washington Post* collected the data from many different sources, there is no way of guaranteeing that the data set is fully comprehensive.

Second, as demonstrated below, all 6,569 victims identified in this data set have been characterized as having a mental illness. As a result, we decide not to make a visualization based on the mental illness variable. We believe this would pose an ethical issue because mental health is a spectrum with varying characteristics and severity of impact. Having a mental illness cannot be reduced to a simple "TRUE" or "FALSE." Instead, we decided that other variables would be a better focus.

```{r}
#| label: mental-illness

sum(police_shootings$`Factors.Mental-Illness`)
```

Finally, a third concern is consent and whether the victims families want their loved ones to have identifying information (like name and information about their death) in a public database. This is a highly sensitive topic and should be treated as such.

We have identified several key variables that will advance our research question: Person.Name - String (Categorical)

`Person.Age` - Integer (Quantitative)

`Person.Gender` - String (Categorical)

`Person.Race` - String (Categorical)

`Incident.Date.Month` - Integer (Categorical)

`Incident.Date.Day` - Integer (Quantitative)

`Incident.Date.Year` - Integer (Quantitative)

`Incident.Date.Full` - String (Categorical)

`Incident.Location.City` - String (Categorical)

`Incident.Location.State` - String (Categorical)

`Factors.Armed` - String (Categorical)

`Factors.Mental-Illness` - Boolean (Categorical)

`Factors.Threat-Level` - String (Categorical)

`Factors.Fleeing` - String (Categorical)

`Shooting.Manner` - String (Categorical)

`Shooting.Body-Camera` - Boolean (Categorical)

**Methodology**

The average age for a police shooting victim is 35; whereas, the median age is 34. We can conclude that the data is not central to 34-35 years of age as the standard deviation of the the victim's age is \~15 years.

```{r}
#| label: summary-stats

police_shootings |>
  summarize(mean_age = mean(Person.Age),
            median_age = median(Person.Age),
            stdv_age = sd(Person.Age))
```

```{r}
police_shootings |>
  mutate(state = Incident.Location.State)
```

```{r}
#install.packages("usmap")
#library(usmap) #import the package
#library(ggplot2) #use ggplot2 to add layer for visualization

#plot_usmap(data = police_shootings, values = "state", regions = "states", color = "red") + 
  #scale_fill_continuous(low = "white", high = "red", name = "Frequency of Police Shootings")
  #labs(title = "U.S. States",
      # subtitle = "This is a blank map of the United States.") 


```

In yearly-analysis, we chose to do a line graph to represent the progression of shooting per day over time as it effectively reveals trends and changes in shooting per day across the years.

```{r}
#| label: yearly-analysis

# Shootings per year
yearly_shootings <- police_shootings |>
  group_by(Incident.Date.Year) |>
  count(Incident.Date.Year) 

days_in_year <- c(365, 366, 365, 365, 365, 366, 253)

yearly_shootings |>
  ggplot(aes(x = Incident.Date.Year, y = n/days_in_year)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Fig. 1: Shootings per Day",
    x = "Year", 
    y = "Number of Shootings")
```

As seen by the graph below, \~10% of male victims and 13% of female victims were unarmed.

```{r}
#| label: arms-status

police_shootings |>
  filter(Factors.Armed != "unknown") |>
  mutate(Factors.Armed.Test = if_else(
    Factors.Armed == "unarmed" | Factors.Armed == "toy weapon", "unarmed", 
    "armed with weapon")) |>
  ggplot(aes(x = Person.Gender, fill = Factors.Armed.Test)) +
  geom_bar(position = "fill") +
  labs(
    title = "Fig. 2: Arms Status of Victims by Gender",
    y = "% of Victims",
    x = "Sex",
    fill = "Arms Status")
```

We used box plots to visualize male versus female and race as it is the best way to show distributions of numeric data values between multiple groups. It made the differences in age between male and female victims and victims of different races very clear. We could compare the median, range, and outliers.

```{r}
#| label: box-plots-race

police_shootings |> 
  ggplot(
    aes(x = Person.Age, color = Person.Race)) + 
    geom_boxplot() + 
    labs(title = "Fig. 3: Relationship between Race and Age",
         x = "Age",
         y = "Race",
         color = "Race")
```

```{r}
#| label: box-plot-gender

police_shootings |> 
  ggplot(
    aes(x = Person.Age, color = Person.Gender)) + 
    geom_boxplot() + 
    labs(title = "Fig. 4: Relationship between Sex and Age",
         x = "Age",
         y = "Sex",
         color = "Sex")
```

```{r}
#| label: bar-graph-race

police_shootings |> 
  ggplot(aes(x = Person.Race, fill = Person.Race)) + 
    geom_bar() + 
    labs(title = "Fig. 5: Number of Victims by Race",
         x = "Race",
         y = "Count",
         fill = "Race") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        legend.position = "none")
```

```{r}
#| label: bar-graph-arms-status-race

police_shootings |>
  filter(Factors.Armed != "unknown") |>
  mutate(Factors.Armed.Test = if_else(
    Factors.Armed == "unarmed" | Factors.Armed == "toy weapon", "unarmed", 
    "armed with weapon")) |>
  ggplot(aes(x = Person.Race, fill = Factors.Armed.Test)) +
  geom_bar(position = "fill") +
  labs(
    title = "Fig. 6: Arms Status of Victims by Race",
    y = "Proportion of Victims",
    x = "Race",
    fill = "Arms Status") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

We will now investigate if the proportion of unarmed people who were shot by police equal for victims of color in comparison to white victims. We hypothesize that the proportion of unarmed people of color shot is significantly greater than the proportion of unarmed white people. In order to investigate this hypothesis, we will perform a hypothesis test using a significance level of $\alpha = 0.05$.

(Note: in the context of this question, we will consider someone as a "POC" if they do not identify as "White")

$\pi_{POC}$ = the proportion of unarmed people out of those shot who identify as a POC.

$\pi_{White}$ = the proportion of unarmed people out of those shot who identify as a White

$H_o$: $\pi_{POC}$ = $\pi_{White}$

The proportion of unarmed people out of those shot who identify as a POC is equal to the proportion of unarmed people out of those shot who identify as white.

$H_A$: $\pi_{POC}$ \> $\pi_{White}$

The proportion of unarmed people out of those shot who identify as a POC is greater than the proportion of unarmed people out of those shot who identify as white.

```{r}
#| label: hypothesis-race

hypo_data <- police_shootings |>
  filter(Factors.Armed != "unknown", Person.Race != "Unknown") |>
  mutate(Factors.Armed.Test = if_else(Factors.Armed == "unarmed" | 
                                        Factors.Armed == "toy weapon", "unarmed"
                                      , "armed with weapon"),
         race = if_else(Person.Race != "White", "POC", "White")) 
set.seed(69)
null_dist <- hypo_data |>
  specify(response = Factors.Armed.Test, 
          explanatory = race , success = "unarmed") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props" , order = c("POC", "White"))
```

```{r}
#| label: hypothesis-race-pvalues
hypo_data |>
  group_by(race, Factors.Armed.Test) |>
  summarize(props = n())

p_hat_POC <- 318/(318+2456)
p_hat_white <- 299/(299+2568)
obs_diff_phat <- p_hat_POC - p_hat_white

visualize(null_dist) +
 shade_p_value(obs_stat = obs_diff_phat, direction = "greater") +
  labs(title = "Fig. 7 Simulated Null Distribution")

null_dist |>
  get_p_value(obs_stat = obs_diff_phat, direction = "greater")

```

**Results**

Based on our summary statistics for the entire data set, we observed that the mean and median age of shooting victims is 35.4 and 34.0 years, respectively. Moreover, the standard deviation of this age range is 14.9 years, demonstrating the wide spread of data points included.

Figure 1 shows a slight decrease in shooting per day in 2021 though the total range of average shootings per day was from roughly 2.45 shootings to 2.8 shootings.

Figure 2 looks at the arms status of victims by gender. This figure shows that female victims were more likely to be armed than male victims. One reason for this is that male victims could have been seen as more aggressive and therefore more likely to become a victim no matter if they were armed.

In Figure 3, we can see that the age distribution of victims does not vary much between races. But if we look at the number of different races in the count of total victims, we can see white victims and African American victims were most common.

As seen in Figure 4, the median age is roughly 34 years old for males and 35 years old for females. Male victims differ by their outliers, as many male victims are over the age of 75. The unknown victims appear to be significantly younger as seen by the median of \~21 years old.

As seen in Figure 5, the most victimized race was white and black. Close to 3000 of the victims were white; whereas, over 1500 victims were black.

As seen in Figure 6, the race with the most unarmed victims was black and the race with the least unarmed victims was Native American. This visualization can provide ample information on implicit bias and how race impacts the police's approaches to interactions with different races.

In Figure 7, we performed a hypothesis test, and got a p-value of 0.107. Utilizing a significance level of 0.05, we fail to reject the null hypothesis that the proportion of unarmed people out of those shot who identify as a POC is equal to the proportion of unarmed people out of those shot who identify as white, as 0.107 is greater than 0.05. Therefore, we have weak evidence to support the alternative hypothesis that the proportion of unarmed people out of those shot who identify as a POC is greater than the proportion of unarmed people out of those shot who identify as white.
